---
title: "Indexes measures for oak decline severity using phenotypic descriptors"
output: pdf_document
keep_tex: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE,cache = TRUE,warning = FALSE,message = FALSE)
```

```{r library_load}
source('../code/library_load.R')
```

```{r prep}
source('../code/data_prep.R')
```

## Abstract

Two indexes, the Phenotypic Decline Index (PDI) and the Decline Acuteness Index (DAI), were developed to describe acute and chronic oak decline severity using a multitude of easily measurable phenotypic descriptors.
Measured phenotypic descriptors include aspects of tree shape and size, crown condition and the presence of biotic agents.
These were collected from 174 trees exhibiting a spectrum of decline severity at 9 sites from across the south, east and west of England.
Inter site differences were identified as a results of aspects of tree size which were corrected for using a mean centring adjustment.
The decline indexes were then calculated using proximity values from an unsupervised random forest of the phenotypic descriptors, multidimensional scaling (MDS) and min-max scaling.
These indexes are simple but sensitive measures of tree decline severity allowing easy comparisons of oak trees both spatially and temporally providing a useful tool for forest monitoring and management.

## Introduction

## Materials and methods

### Phenotypic data collection

### Phenotypic descriptor preparation and analysis

## Results

### Phenotypic descriptors

The following four additional descriptors were calculated:

**Crown condition (%)** - percentage of present crown proportional to the percentage crown density.

$$c = (100 - m)(1-\frac{t}{100})$$

$m$ = missing crown (%), $t$ = crown transparency (%)

**Estimated crown volume (m^3^)** - estimated volume of crown as an ellipsoid proportional to the crown condition.

$$v = \frac{c(h - l)}{150}\pi r^2$$

$r$ = crown radius (m), $h$ = total height (m), $l$ = lower crown height (m), $c$ = crown condition (%)

**Estimated bleed prevalence (%)** - estimated percentage of the surveyed trunk area affected by bleeds or black staining.

$$p = \frac{a^2A + b^2B}{3d\pi}$$

$a$ = average active bleed size (cm), $A$ = number of active bleeds, $b$ = average black stain size (cm), $B$ = black stain number, $d$ = diameter at breast height (cm)

***Agrilus* exit hole density (m^-2^)** - density of observed *Agrilus* exit holes across the surveyed trunk area.
$$e = \frac{h}{\frac{2d}{100}\pi}$$
$h$ = number of *Agrilus* exit holes, $d$ = diameter at breast height (cm)

### Tree size descriptor adjustment by site

### Phenotypic decline and decline acuteness indexes

## Discussion

### Applications

## Conclusions

## Figures

```{r figures,child='figures.Rmd'}
```

## Checks

```{r check,comment=''}
source('../code/check.R')
manuscript_check()
```

## References