---
title: "Index measures for oak decline severity using phenotypic descriptors"
output: 
  bookdown::pdf_document2:
   toc: false
   number_sections: false
   keep_tex: no
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE,
                      warning = FALSE,
                      message = FALSE)
```

```{r loadData, include = FALSE}
loadd(site_rf_margin)
loadd(site_rf_post_correction_margin)
```

**Authors:** Jasen Finch, Nathan Brown, Sandra Denman & John Draper

**Proposed journal:** Still undecided! ??Plant Methods, Forestry?? 

**Keywords:** acute oak decline, chronic oak decline, phenotyping, machine learning

## Abstract

Two indexes, the Phenotypic Decline Index (PDI) and the Decline Acuteness Index (DAI), were developed to describe acute and chronic oak decline severity using a multitude of easily measurable phenotypic descriptors.
Measured phenotypic descriptors include aspects of tree shape and size, crown condition and the presence of biotic agents.
These were collected from 174 trees exhibiting a spectrum of decline severity at 9 sites from across the south, east and west of England.
Inter site differences were identified as a results of aspects of tree size which were corrected for using a mean centring adjustment.
The decline indexes were then calculated using proximity values from an unsupervised random forest of the phenotypic descriptors, multidimensional scaling (MDS) and min-max scaling.
These indexes are simple but sensitive measures of tree decline severity allowing easy comparisons of oak trees both spatially and temporally providing a useful tool for forest monitoring and management.

## Introduction

- define phenotype and descriptor
- Acute and chronic decline associated phenotypes
- The oak decline spectrum move away from simplistic binary categorisations
- Manual assignment of decline status difficult prone to error for due to contribution of many factors. 

Oak decline syndromes are facilitated by a variety of both abiotic and biotic factors.

Abiotic factors can include...
Biotic factors can include...

As a results of this, there are many visual cues measures that contribute to what can be defined as a declining oak tree.
A severely declining oak is prominantly characterised by very poor crown condition and numerous dead branches.
Compounding this complexity is the addition of different types of oak decline (AOD) that are often categorised based on the duration over which a tree has been declining or the biotic agents/factors associated with a declining tree.
For instance a chronically declining oak can subsequently begin to also develop the stem bleeds associated with AOD.
Accurately diagnosing in the field the severity and type of oak decline present is difficult and can requires much experience, expertise and training.

Here, we propose a framework for quantitatively measuring the severity of oak decline 

* The phenotypic decline index (PDI) - a measure of decline severity scoring trees between 0 and 1.
More severly declining oak trees have a score closer to 1.

* The decline acuteness index (DAI) - a measure differentiating between chronically and acutely declining oak trees with a score between -1 and 1. 
Acutely declining trees have a score closer to 1 and chronically declining trees have a score closer to -1.


## Materials and methods

### Phenotypic data collection

- Site descriptions
- Field phenotyping protocol

Phenotypic data were collected from a total of 174 trees from 9 sites from across the south, eas and west of England.


### Phenotypic descriptor preparation and analysis

- Aggregation of cardinal observations.
- Random forest and MDS analyses

R v`r str_c(version$major,version$minor,sep ='.')` was used for all parsing, preparation and analysis of phenotypic data.

## Results

### Phenotypic descriptors for oak decline

- Descriptions of collected phenotypic descriptors and manual status assignment

The following four additional descriptors were calculated:

**Crown condition (%)** - percentage of present crown proportional to the percentage crown density.

$$c = (100 - m)(1-\frac{t}{100})$$

$m$ = missing crown (%), $t$ = crown transparency (%)

**Estimated crown volume (m^3^)** - estimated volume of crown as an ellipsoid proportional to the crown condition.

$$v = \frac{c(h - l)}{150}\pi r^2$$

$r$ = crown radius (m), $h$ = total height (m), $l$ = lower crown height (m), $c$ = crown condition (%)

**Estimated bleed prevalence (%)** - estimated percentage of the surveyed trunk area affected by bleeds or black staining.

$$p = \frac{a^2A + b^2B}{3d\pi}$$

$a$ = average active bleed size (cm), $A$ = number of active bleeds, $b$ = average black stain size (cm), $B$ = black stain number, $d$ = diameter at breast height (cm)

***Agrilus* exit hole density (m^-2^)** - density of observed *Agrilus* exit holes across the surveyed trunk area.
$$e = \frac{h}{\frac{2d}{100}\pi}$$
$h$ = number of *Agrilus* exit holes, $d$ = diameter at breast height (cm)

### Tree size descriptor adjustment by site

- Analysis of cross site differences
- Outline of cross site correction approach using mean centering adjustment

Large phenotypic differences were found between the sites by supervised random forest (Figure #a; margin = `r site_rf_margin$Margin %>% signif(digits = 3)`). 
This was as a results of differences in descriptors relating tree size and likely the ages of the trees between the sites. 
To account for these differences and to make the trees across the sites comparable in the context of decline, an adjustment of site means to the overall mean was applied to each of the size descriptors (diameter at breast height, lower crown height, timber height, total height, crown radius). 
An example of this adjustment is shown for diameter at breast height in Supplementary Figure # and the adjustment factors for each of the descriptors at each site is shown in Supplementary Table #.  
This adjustment removed inter-site variability but maintained intra-site variability, allowing comparisons to be made between the trees at the different sites in the context of decline severity.
Post adjustment the random forest margin was substantially reduced to `r site_rf_post_correction_margin$Margin %>% signif(digits = 3)` indicating that the majority of the between site variance had been removed. 

### Phenotypic decline and decline acuteness indexes for oak decline severity

- Decline index calculation
- Comparison of indexes with manually assigned decline status labels
- Top ranked descriptor index response surfaces

Using unsupervised random forest proximities and the phenotypic descriptors as reported previously,multidimensional scaling (MDS) to two dimensions was performed to yield two phenotypic indexes:

* **Phenotypic decline index (PDI)** - the inverse of the phenotypic health index reported previously providing a measure of how severely a tree is declining. 
Min-max scaling performed on the first MDS dimension giving score between 0 and 1, with more severely declining trees having a score closer to 1.

* **Decline acuteness index (DAI)** - a measure that allows the differentiation between chronically and acutely declining oaks. 
The second MDS dimension was min-max scaled to give scores between -1 and 1 with chronically declining trees having a score closer to -1 and acutely declining trees having a score closer to 1. 
Trees with a score around 0 are neither distinctively AOD or COD sharing characteristics of both conditions in trees with PDI greater than 0.3.

The two dimensional space provided by these phenotypic indexes, for measuring decline severity isshown in Figure 1. 
The phenotypic descriptors used to calculate these indexes are easily measurable in the field and require no specialist equipment. 
They allow sensitive comparisons to be made for the same tree between years as well as comparisons of trees between different locations, which could have great utility in long term forest health monitoring for both forestry management and research.

Crown related descriptors were found to contribute most to the PDI (Figure 2a) with more severely declining trees having reduced crown volume and condition. 
The DAI was more complex in descriptor contribution with trunk staining and bleeding descriptors contributing highly with bleed prevalence contributing the most. Trees with any black staining or bleeding had scores greater than 0. 
Trees with high bleed prevalence and therefore a larger area of their trunk covered with black staining or active bleeds had much higher DAI scores.

The phenotypic indexes of all 171 trees (including the 20 trees Chestnuts) are shown in Figure 3. 
It can be seen that across the 9 sites, purely chronic oak decline sites such as Chestnuts and Speculation are easily distinguishable using the DAI compared to more AOD dominant sites such as Langdale or mixed sites such as Great Monks Wood.

This plot highlights the inaccuracies in manual field classifications of decline status with several trees classified as healthy controls that, when compared quantitatively to other trees, are in fact
in a decline state. 
Most notably this has occurred in a number of trees at Attingham, Hatchlands and Winding Wood. 
Also notable from Figure 3 is the disparity between the relative health of the healthier trees across the sites. 
The healthier trees at sites such as Attingham and Hatchlands were found to be in a more declined state than those at sites such as Langdale and Chestnuts.

### Trends of key phenotypic descriptors within the PDI and DAI measures

The Phenotypic Decline Index (PDI) and Decline Acuteness Index (DAI) were previously defined in the 30 month report. 
A machine learning approach was used to derive these measures as it allows the complex interactions and trends between phenotypic descriptors to be modelled from which the decline indexes are calculated. 
This would likely be oversimplified using conventional modelling strategies. 
These complex interactions can be investigated by simulating phenotypic data and generating the index values from random forest models that have been trained for each of the indexes, supervised using phenotypic descriptors from the 171 trees. 
The index values generated by the models from these simulated phenotypic data can then be visualised as response surface plots, from which descriptor interactions within the indexes can be interpreted. 
Figures 1 and 2 show the simulated index value response surfaces under a variety of different circumstances for key phenotypic descriptors that contribute highly to the PDI and DAI.

#### PDI

Crown condition descriptors such as crown volume and crown condition contributed highly to the PDI. 
The phenotypic measures central to this were percentage missing crown and percentage crowntransparency from which crown condition and crown volume are calculated. 
It can be seen in Figure 1 that as the percentage missing crown and percentage crown transparency increase and therefore crown condition and estimated crown volume decrease, the PDI increases. 
A much greater range of PDI values can be obtained from trees with relatively greater stature (e.g. greater crown radius, total height, lower crown height and diameter at breast height; Figures 1a & b), with relatively small stature trees having PDIs of 0.5 or greater even with 100% crown condition.

#### DAI

The DAI is a much more complex index with more descriptors contributing to it than the PDI such as bleed prevalence, diameter at breast height and crown condition descriptors. 
Simulated response surfaces for the DAI are shown in Figure 2. 
These response surfaces show circumstances under which bleed prevalence related descriptors (Figures 2a, b & c) and crown condition descriptors (Figures 2d, e & f) change for non-symptomatic AOD or COD trees (Figures 2a & d respectively), moderate AOD or COD (Figures 2b & e respectively) and severe AOD or COD (Figures 2c & f respectively). 
Increasing active bleed, staining size and frequency contributed to greater positive values of the DAI and more severe AOD. 
Also higher DAI values could be obtained from trees with greater stature (Figure 2c). 
Similarly to the PDI, trees with smaller stature and poorer crown condition give lower DAI values (Figures 2d, e & f) and therefore more severe COD.
The relationship of relative tree stature and the acuteness of the decline likely reflects the period of time under which these trees have been afflict. 
AOD occurs over a much shorter period compared COD and therefore has less time in which to inhibit growth of the tree as a whole.

## Discussion

- Accessability of collecting descriptors
- Suitability of collected phenotypic descriptors for describing decline status
- Concession of model interpretability to predictive capability by using a machine learning approach
- Trends of key phenotypic descriptors withing the PDI and DAI mesures
- Sensitivity provided for integration with large scale omics analyses as well as long term monitoring for forest condition
- Index thresholds for category definitions (healthy, declining, severe decline etc.)
- When viewed in the 2D continuum, the categorisation of AOD to only include bleeds of certain bacteria seems likely to be too specific and by definition should only describe the period of time under which the tree have been declining.
Perhaps subcategories of AOD would be more suitable such as Agrillus associated AOD or Armillaria associated AOD. 

### Applications

- Discussion of applications to forestry monitoring and management
- Sensitivity for integration with molecular data compared to categorical assignments

## Conclusions

- Crown condition descriptors most important for describing overall decline severity in the PDI.
- Bacterial bleed related descriptors are most important for describing decline acuteness.
- Decline indexes provide intuitive measures of decline severity based on simple phenotypic measurements that do not require specialist equipment

\newpage
## Checks

```{r check,comment='',cache=FALSE}
manuscript_check()
```

## References